A generalized quadratic estimate for random field nonstationarity \thanksref{code}

### 2016-07-01 paper/paper.tex
added notes to the manuscript in reference to what was discussed on skype

The seminal work of Dahlhaus in the time series literature \cite{dahlhaus1997fitting,dahlhaus2000likelihood}  is  an example of a collection of results that hint at a more general statistical theory of nonstationarity.  Dahlhaus develops asymptotic theory for a particular class of nonstationary random fields modeled by a spatially (or temporally) varying spectral density. These random fields were originally developed for time series \cite{priestley1965evolutionary,priestley1981spectral} and have the form 
\begin{align}
\label{eq: intro nonstat phase model}
Z(\bs x)= \int_{\Bbb R^d} e^{i\bs x\cdot \bs k} A(\bs k, \bs x)\sqrt{C_{\bs k}}\frac{dW_{\bs k}}{(2\pi)^{d/2}}
\end{align} 
where  $C_{\bs k}$ is a spectral density, $dW_{\bs k}$ is an orthogonal increment random complex measure that satisfies $E|dW_{\bs k}|^2 =  d{\bs k}$ (see \cite{Gikhman_Skorokhod_book_v1} for details on random measures) and $A({\bs k},{\bs x})$ represents a spatial (or temporal) modulation of $\sqrt{C_{\bs k}}$. 
Dahlhaus proved consistency results for estimating the squared modulus $|A({\bs k},{\bs x})|^2$ when estimation is done  by maximizing a weighted sum of local likelihood functions. Left unanswered, however, is the question of estimating the phase of $A({\bs k},{\bs x})$. In Section \ref{section: NPhase}, we partly resolve this question by showing that a generalized quadratic estimate can be used to estimate a pure phase modulation $A({\bs k},{\bs x})=\exp({iB(\bs k,\bs x)})$ where $B(\bs k,\bs x)$ is a function taking values in $\Bbb R$ and is a separable function of $\bs x$ and $\bs k$. Indeed, this nonstationary phase model has the local invariance property and, as such, can be accurately estimated (under certain conditions on $B(\bs k,\bs x)$) with the generalized quadratic estimate developed here.

@@

@@

Based on assumptions \ref{Assumption 4}, \ref{Assumption 2} and \ref{Assumption 3} given in the previous section, the first step for deriving a generalized quadratic estimate of $\phi(\bs x)$ is to linearly approximate  $C^{\bs \theta}(\bs x,\bs y)$, with a Taylor expansion in $\bs \theta(\bs x) - \bs \theta(\bs y)$ as follows 
\begin{align}
\label{first order term, intro}
C^{\bs \theta}(\bs x,\bs y) = C^{(0)}(\bs x-\bs y)
        + \bs C^{(1)}(\bs x-\bs y)\cdot \left({\bs\theta(\bs x)-\bs\theta(\bs y)}\right)
        + \mathcal O({\phi}^2)+ \mathcal O(\phi^3)+\cdots.
\end{align}
where $C^{(0)}\colon \Bbb R^d \rightarrow \Bbb R$ and $\bs C^{(1)}\colon \Bbb R^d \rightarrow \Bbb R^d$ satisfies $\bs C^{(1)}(-\bs x) = -\bs C^{(1)}(\bs x)$.
Now, truncating (\ref{first order term, intro}) to first order and applying Claim \ref{claim: first order expansion} of Appendix \ref{section: Detailed Proofs} gives the following linear approximation of the cross frequency covariance in the Fourier transform of $Z(\bs x)$
\begin{align}
\label{eq: first order cov intro}
E\big(Z_{\bs k+\bs \ell}Z_{-\bs k}\big)  \approx  \phi_{\bs\ell}\left({\bs\xi_{\bs\ell}\!\cdot\! \bs C^{(1)}_{\bs k} - \bs\xi_{\bs\ell}\!\cdot\! \bs C_{\bs k+\bs \ell}^{(1)}}\right).
\end{align}
Recall Bochner's Theorem (details can be found in \cite{Gikhman_Skorokhod_book_v1}) which states that the random field $Z(\bs x)$ is stationary if and only if  \mbox{$E\big(Z_{\bs k+\bs \ell}Z_{-\bs k}\big) = 0$} for every $\bs\ell \neq\bs 0$. Therefore nontrivial covariance between $Z_{\bs k+\bs \ell}$ and $Z_{-\bs k}$ at some nonzero lag $\bs\ell$ provides a direct probe into the nonstationarity present in $Z(\bs x)$. Equation (\ref{eq: first order cov intro}) is, therefore, a translation of how local invariant nonstationarity relates to nonzero cross covariance in $Z_{\bs k}$.
Using this translation, Claim \ref{appendix, claim: quad est derivations} of Appendix \ref{section: Detailed Proofs} derives the quadratic estimate of $\phi_{\bs \ell}$. This estimate is effectively  an inverse-variance weighted autocovariance estimate in the Fourier domain and is given by
\begin{align}
    \label{explicit form of the qe in body of paper}
    \hat\phi_{\bs \ell}
    &=  A_{\bs \ell}
        \int
        {\Big(\bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k} - \bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k+\bs \ell}\Big)}^{\! *}
        \frac{Z^{obs}_{\bs k+\bs \ell}Z^{obs}_{-\bs k}}{C^{ZZobs}_{\bs k+\bs \ell}C^{ZZobs}_{\bs k}}\frac{d\bs k}{{(2\pi)}^{d/2}}.
\end{align}
In the above formula, $A_{\bs \ell}$ is a normalizing constant (see Claim \ref{appendix, claim: quad est derivations} for an exact expression),  $C^{ZZobs}_{\bs k}$ is defined to be $C^{ZZm}_{\bs k} + C^{N\!N}_{\bs k}$ where $C^{ZZm}_{\bs k}$ is the spectral density of $Z(\bs x)$ marginalized over the prior for $\phi(\bs x)$ given in Assumption \ref{Assumption 3}. Notice that the prior only serves to optimize the weights in $\hat\phi_{\bs \ell}$. Indeed, one can easily avoid specifying $C_{\bs k}^{\phi\phi}$ by instead defining $C^{ZZm}_{\bs k}$ to be ${(2\pi)}^{d/2}C^{(0)}_{\bs k}$ where $C^{(0)}_{\bs k}$ denotes the Fourier transform of $C^{(0)}(\bs x)$ from (\ref{first order term, intro}).

@@

>>>>

The seminal work of Dahlhaus in the time series literature \cite{dahlhaus1997fitting,dahlhaus2000likelihood}  is  an example of a collection of results that hint at a more general statistical theory of nonstationarity.  Dahlhaus develops asymptotic theory for a particular class of nonstationary random fields modeled by a spatially (or temporally) varying spectral density. These random fields were originally developed for time series \cite{priestley1965evolutionary,priestley1981spectral} and have the form
\begin{align}
\label{eq: intro nonstat phase model}
Z(\bs x)= \int_{\Bbb R^d} e^{i\bs x\cdot \bs k} A(\bs k, \bs x)\sqrt{C_{\bs k}}\frac{dW_{\bs k}}{(2\pi)^{d/2}}
\end{align}
where  $C_{\bs k}$ is a spectral density, $dW_{\bs k}$ is an orthogonal increment random complex measure that satisfies $E|dW_{\bs k}|^2 =  d{\bs k}$ (see \cite{Gikhman_Skorokhod_book_v1} for details on random measures) and $A({\bs k},{\bs x})$ represents a spatial (or temporal) modulation of $\sqrt{C_{\bs k}}$.
Dahlhaus proved consistency results for estimating the squared modulus $|A({\bs k},{\bs x})|^2$ when estimation is done  by maximizing a weighted sum of local likelihood functions. Left unanswered, however, is the question of estimating the phase of $A({\bs k},{\bs x})$. In Section \ref{section: NPhase}, we partly resolve this question by showing that a generalized quadratic estimate can be used to estimate a pure phase modulation $A({\bs k},{\bs x})=\exp({iB(\bs k,\bs x)})$ where $B(\bs k,\bs x)$ is a function taking values in $\Bbb R$ and is a separable function of $\bs x$ and $\bs k$. Indeed, this nonstationary phase model has the local invariance property and, as such, can be accurately estimated (under certain conditions on $B(\bs k,\bs x)$) with the generalized quadratic estimate developed here.
\textcolor{red}{
Try and add this to the intro... emphasize that the form of the local invariant nonstationarity gives this quasi-stationary form in the Fourier domain which allows FFT methods for empirical cov estimation.
\begin{align}
E\big(Z_{\bs k+\bs \ell}Z_{-\bs k}\big)  \approx  \phi_{\bs\ell}\left({\bs\xi_{\bs\ell}\!\cdot\! \bs C^{(1)}_{\bs k} - \bs\xi_{\bs\ell}\!\cdot\! \bs C_{\bs k+\bs \ell}^{(1)}}\right).
\end{align}
}

@@

\textcolor{red}{Be explicit that all function defined on $\Bbb R^d$ are periodic and the integrals over real space are restricted to one periodic region etc...}

@@

\textcolor{red}{Consider giving the estimate up front, along with the FFT represention, then explain where it comes from.}

Based on assumptions \ref{Assumption 4}, \ref{Assumption 2} and \ref{Assumption 3} given in the previous section, the first step for deriving a generalized quadratic estimate of $\phi(\bs x)$ is to linearly approximate  $C^{\bs \theta}(\bs x,\bs y)$, with a Taylor expansion in $\bs \theta(\bs x) - \bs \theta(\bs y)$ as follows
\begin{align}
\label{first order term, intro}
C^{\bs \theta}(\bs x,\bs y) = C^{(0)}(\bs x-\bs y)
        + \bs C^{(1)}(\bs x-\bs y)\cdot \left({\bs\theta(\bs x)-\bs\theta(\bs y)}\right)
        + \mathcal O({\phi}^2)+ \mathcal O(\phi^3)+\cdots.
\end{align}
where $C^{(0)}\colon \Bbb R^d \rightarrow \Bbb R$ and $\bs C^{(1)}\colon \Bbb R^d \rightarrow \Bbb R^d$ satisfies $\bs C^{(1)}(-\bs x) = -\bs C^{(1)}(\bs x)$.
\textcolor{red}{Consider giving the explicit derivative form of $\bs C^{(1)}(\bs x-\bs y)$.}
Now, truncating (\ref{first order term, intro}) to first order and applying Claim \ref{claim: first order expansion} of Appendix \ref{section: Detailed Proofs} gives the following linear approximation of the cross frequency covariance in the Fourier transform of $Z(\bs x)$
\begin{align}
\label{eq: first order cov intro}
E\big(Z_{\bs k+\bs \ell}Z_{-\bs k}\big)  \approx  \phi_{\bs\ell}\left({\bs\xi_{\bs\ell}\!\cdot\! \bs C^{(1)}_{\bs k} - \bs\xi_{\bs\ell}\!\cdot\! \bs C_{\bs k+\bs \ell}^{(1)}}\right).
\end{align}
Recall Bochner's Theorem (details can be found in \cite{Gikhman_Skorokhod_book_v1}) which states that the random field $Z(\bs x)$ is stationary if and only if  \mbox{$E\big(Z_{\bs k+\bs \ell}Z_{-\bs k}\big) = 0$} for every $\bs\ell \neq\bs 0$. Therefore nontrivial covariance between $Z_{\bs k+\bs \ell}$ and $Z_{-\bs k}$ at some nonzero lag $\bs\ell$ provides a direct probe into the nonstationarity present in $Z(\bs x)$. Equation (\ref{eq: first order cov intro}) is, therefore, a translation of how local invariant nonstationarity relates to nonzero cross covariance in $Z_{\bs k}$.
Using this translation, Claim \ref{appendix, claim: quad est derivations} of Appendix \ref{section: Detailed Proofs} derives the quadratic estimate of $\phi_{\bs \ell}$. This estimate is effectively  an inverse-variance weighted autocovariance estimate in the Fourier domain and is given by
\begin{align}
    \label{explicit form of the qe in body of paper}
    \hat\phi_{\bs \ell}
    &=  A_{\bs \ell}
        \int
        {\Big(\bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k} - \bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k+\bs \ell}\Big)}^{\! *}
        \frac{Z^{obs}_{\bs k+\bs \ell}Z^{obs}_{-\bs k}}{C^{ZZobs}_{\bs k+\bs \ell}C^{ZZobs}_{\bs k}}\frac{d\bs k}{{(2\pi)}^{d/2}}.
\end{align}
In the above formula, $A_{\bs \ell}$ is a normalizing constant (see Claim \ref{appendix, claim: quad est derivations} for an exact expression),  $C^{ZZobs}_{\bs k}$ is defined to be $C^{ZZm}_{\bs k} + C^{N\!N}_{\bs k}$ where $C^{ZZm}_{\bs k}$ is the spectral density of $Z(\bs x)$ marginalized over the prior for $\phi(\bs x)$ given in Assumption \ref{Assumption 3}. Notice that the prior only serves to optimize the weights in $\hat\phi_{\bs \ell}$. Indeed, one can easily avoid specifying $C_{\bs k}^{\phi\phi}$ by instead defining $C^{ZZm}_{\bs k}$ to be ${(2\pi)}^{d/2}C^{(0)}_{\bs k}$ where $C^{(0)}_{\bs k}$ denotes the Fourier transform of $C^{(0)}(\bs x)$ from (\ref{first order term, intro}).

@@

\textcolor{red}{Re-write this part to be explicit about what is rigourous and what is not. Maybe consider moving the rigourous algebraic manipuliations relating the Fourier characterization and the fast algorithm to a claim in the main body of the paper.}
\textcolor{magenta}{Consider actually proving (or citing) that the periodic transform follows the same behavior of the continuous trasform, whereby justifying the notational convention.}
